## Autism {#caseStudyAutism}

```{r}

kids.usernames <- c("fbmgb","preston","fbmgg")
kids <- subset_samples(people.phylum, Username %in% kids.usernames)
kids.phylum <- prune_taxa(taxa_sums(kids)>42,kids)

kids.genus <- subset_samples(subset_samples(people.norm, Username %in% kids.usernames))
kids.genus <- prune_taxa(taxa_sums(kids.genus)>42,kids.genus)

kids.species <- subset_samples(subset_samples(people.species, Username %in% kids.usernames))
kids.species <- prune_taxa(taxa_sums(kids.species)>42,kids.species)

autism.phylum <- subset_samples(kids.phylum, Username == "fbmgb")
autism.genus <- subset_samples(kids.genus, Username == "fbmgb")
boys.phylum <- subset_samples(kids.phylum, Username == "preston")
boys.genus <- subset_samples(kids.genus, Username == "preston")
```

What can we learn from the microbiome of a 4-year-old boy with Autism? A mother sent me the Explorer results of her 4-year-old, who suffers from Autistic Spectrum Disorder (ASD)

Here's the overall picture of the phlyum-level microbes in his gut:

```{r}
kable(mhg_abundance(autism.phylum)/10000, digits = 2, col.names = "ASD")
```

Looks quite normal, especially for somebody on an omnivore diet. Lots of Firmicutes and reasonable Actinobacteria. Proteobacteria is a smidge high, though not unusual for a single sample.

Let's look at more details, down to the genus level, and compare him to some similarly-aged (healthy) kids:

```{r}

kids.table <- mhg_abundance(kids.genus, colnames = "Label")/10000
kable(kids.table[order(kids.table[,2], decreasing = TRUE),], digits = 2, col.names = c("Girl","ASD","Boy"))#sample_data(kids.genus)$Username)
```

Now, what about the autistic boy is unique?  Let's compare him to a similar-aged, healthy boy and see which microbes are present in the autistic case but _not_ the healthy case:

```{r}
kable(psmr::unique_taxa(subset_samples(kids.genus, Username == "fbmgb"), subset_samples(kids.genus, Username == "preston"))[1]/10000, digits = 2, col.names = "ASD")
```

The autistic boy has _Sutterella_, an organism that is missing in the healthy boy.

How unusual is that level of _Sutterella_?  I don't have enough samples of healthy boys to do a real comparison, but just to get a rough idea of what we're dealing with, here's the range over a mix of 100+ health and unhealthy samples (mostly adults).  The red line indicates where the autistic boy fits on the range of _Sutterella_ abundances:

```{r}


ggplot(data=data.frame(abundance=as.numeric(otu_table(people.norm)["Sutterella"])/10000,
  label = sample_data(people.norm)$Username,
  date = sample_data(people.norm)$Date),
  aes(x=abundance)) + 
 stat_density() +
  annotate(geom="segment", color = "red",
           x = as.numeric(otu_table(subset_samples(kids.genus, Username == "fbmgb"))["Sutterella"])/10000,
           xend = as.numeric(otu_table(subset_samples(kids.genus, Username == "fbmgb"))["Sutterella"])/10000 ,
           y = -Inf, yend = Inf)


```


#### Microbes of known association with autism

_Clostridium tetani_ may play a role in autism [ citation needed ], though unfortunately our 16S test can't see this microbe. Instead let's look for that microbe at the genus level:



```{r}


ggplot(data=data.frame(abundance=as.numeric(otu_table(people.norm)["Clostridium"])/10000,
  label = sample_data(people.norm)$Username,
  date = sample_data(people.norm)$Date),
  aes(x=abundance)) + 
 stat_density() +
  annotate(geom="segment", color = "red",
           x = as.numeric(otu_table(subset_samples(kids.genus, Username == "fbmgb"))["Clostridium"])/10000,
           xend = as.numeric(otu_table(subset_samples(kids.genus, Username == "fbmgb"))["Clostridium"])/10000 ,
           y = -Inf, yend = Inf)


```

Although obviously a single test can't say much, he is not especially unusual for this microbe, at least at the genus level.

