## Other People {#CaseStudyOther}

Here's a person from UK who is out of range on two specific microbes:

```{r caseStudysh1}

psmr::mhg_outlier(psmr::mhg_range_for(subset_samples(people.norm, Site == "gut" & Username == "sh1")),
                 psmr::mhg_range_for(people.healthy.gut.genus)) %>% knitr::kable()

# Subject has been taking Miyarisan probiotic supplements (Clostridium butyricum).  Does it show?

s <- subset_samples(people.norm, Site == "gut" & Username == "sh1")
#psmr::mhg_taxa(people.species,"Clostridium butyricum")  # none
```

Is this good or bad? I don't know.


```{r caseStudySB}

sb.phylum <- subset_samples(people.gut.phylum.norm, Username == "sb")
sb.norm <- subset_samples(people.gut.norm, Username == "sb")

tam.phylum <- subset_samples(people.gut.phylum.norm, Username == "tm")
tam.norm <- subset_samples(people.gut.norm, Username == "tm")

# mhg_abundance(tam.norm)/10000
# mhg_abundance(tam.phylum)/10000

#cb.json <- jsonlite::read_json(file.path(path_to_user("Charles_Bergmeier"),"Charles_Bergmeier-gut-2016-04-06.json"))

dr.phylum <- subset_samples(people.gut.phylum.norm, Username == "dr2")
dr.norm <- subset_samples(people.gut.norm, Username == "dr2")

# mhg_abundance(dr.phylum)/10000
# 
# mhg_unique(tam.norm, dr.norm)
# 
# mhg_taxa(tam.norm, "Bifidobacterium")
# mhg_taxa(tam.norm, "Lactobacillus")

# df <-   # mhg_all_who_have(people.norm,"Bifidobacterium") %>% filter(Condition=="Healthy" & Username!="tt") %>% arrange(desc(Abundance))
# 
# 
# 
# ggplot(data=df , aes(x=reorder(Username,-Abundance),y=Abundance)) + geom_bar(stat="identity") +
#   theme(axis.text.x = element_text(angle=90)) +
#   labs(title="Abundance of Bifidobacteria in hundreds of healthy samples")



# mhg_abundance(sb.phylum)/10000
# mhg_abundance(sb.norm)/10000

make_ord_for(subset_samples(people.gut.norm,Age<30 & Age>18))

```

ub7 complains of dysbiosis. ub16 is a very healthy, active female.  bp1 began a ketogenic diet.

```{r caseStudySB2}
# 
# comparables <- subset_samples(people.gut.norm,Age<30 & Age>18 & !Username=="bp1" | (Username == "bp1" & Label == "Before Diet"))
# 
# mhg_plot_top_heat(comparables, n = 10)
# mhg_abundance(comparables)/10000
# 
mhg_density_for_taxa(merge_phyloseq(gut.norm,people.norm), "Blautia") +
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "tm")
  )["Blautia"]) / 10000,
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "tm")
  )["Blautia"]) / 10000 ,
  y = -Inf,
  yend = Inf
  )

# otu_table(gut.norm)[,otu_table(gut.norm)["Alistipes"]>40000]
# 
# otu_table(gut.norm)["Alistipes"]>40000
# 
# mhg_density_for_taxa(gut.norm,"Blautia")

```

How about another kefir drinker?


```{r caseStudyOtherKefirDri}
lm.norm <- subset_samples(people.norm, Username == "lm")
lm.phylum <- subset_samples(people.phylum, Username == "lm")
mhg_abundance(lm.phylum)/10000
mhg_abundance(lm.norm, colnames="Date")/10000
mhg_density_for_taxa(merge_phyloseq(people.norm,gut.norm), "Anaeroplasma") +
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(
  subset_samples(people.norm, Username == "lm")
  )["Anaeroplasma"]) / 10000,
  xend = as.numeric(otu_table(
  subset_samples(people.norm, Username == "lm")
  )["Anaeroplasma"]) / 10000 ,
  y = -Inf,
  yend = Inf
  )

#mhg_taxa(subset_samples(people.norm, Site == "gut"),"Anaeroplasma")

#sample_data(subset_samples(people.norm, SSR %in% colnames(otu_table(people.norm)["Anaeroplasma",otu_table(people.norm)["Anaeroplasma"]>0])))[,c("Username","Date")]

df <- mhg_all_who_have(people.norm,"Fusicatenibacter") %>% arrange(desc(Abundance))



ggplot(data=df , aes(x=reorder(Label,-Abundance),y=Abundance)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title="Abundance of Fusicatenibacter in hundreds of samples") 
#  annotate(geom = "point", x = df %>% filter(Username=="pra") %>% select(Label), y = -Inf)


```

And a family of Australians

```{r caseStudyOtherAussie1}
aus.phylum <- subset_samples(people.phylum, grepl("lm[a-z]*",Username))
aus.norm <- subset_samples(people.norm, grepl("lm[a-z]*",Username))
aus.wife <- subset_samples(aus.norm, SSR == 445566)

# people.tmp <- subset_samples(people.norm, Site=="gut" & Condition %in% c("Healthy"))
# sample_data(people.tmp)$Username <- "Other"
# sample_data(people.tmp)$Username[sample_data(people.tmp)$SSR %in% sample_data(aus.norm)$SSR] <- "AUS"
# make_ord_for(people.tmp)

# husband 445587
# son 447247
sample_data(aus.norm)$Label <- c("son","father","mother")
#mhg_abundance(aus.phylum, colnames="Label")/10000
mhg_plot_top_heat(aus.norm)
```


Here are the taxa that are found in unusual high abundances compared to healthy people
```{r caseStudyOtherAussieHealthyRanges}

# all taxa where aus.norm is greater than healthy people's max
mhg_outlier(mhg_range_for(aus.norm),mhg_range_for(subset_samples(people.norm, Site=="gut" & Condition=="Healthy"))) %>% filter(original>max) %>% select(taxa,mean,sd,original) %>% knitr::kable(digits=2, caption = "This sample is out of healthy range for these taxa.")



```

Let's look for _Anaeroplasma_

```{r caseStudyOtherAussie}

ps <- people.norm
# sample_data(ps)$Username <- "P"
#grepl("lm[a-z]*",
      

#sample_data(ps)[,"Username"] == "lm"
ggplot(data= mhg_all_who_have(ps,"Anaeroplasma") %>% arrange(desc(Abundance)) , aes(x=reorder(Label,-Abundance),y=Abundance)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title="Abundance of Anaeroplasma in hundreds of samples",
       y = "Abundance (%)",
       x = "Person")

#sample_data(people.norm)[,c("Username","Date","Label")] %>% arrange(Label)
```



Another user, ```sj```, here showing all the taxa that are out of range of healthy people (Figure: \@ref(tab:caseStudyOtherSJRange))

```{r caseStudyOtherSJRange}

sj <- subset_samples(people.norm,Username == "sj" & Site == "gut")
sj.phylum <- subset_samples(people.phylum, Username == "sj" & Site == "gut" )
sj.family <- subset_samples(people.family, Username == "sj" & Site == "gut")
ps <- subset_samples(people.gut.norm, Site == "gut" )
users <- sample_data(ps)$Username
users[-which(users=="sj")] <- "Person"
sample_data(ps)$Username <- users

#make_ord_for(merge_phyloseq(subset_samples(ps,Condition=="Healthy"),sj)) + labs(title="Compared to Healthy Users")
#mhg_abundance(sj,top = 50)/10000

# all taxa where sj is greater than healthy people's max
mhg_outlier(mhg_range_for(sj),mhg_range_for(subset_samples(people.norm, Site=="gut" & Condition=="Healthy"))) %>% filter(original>max) %>% select(taxa,mean,sd,original) %>% knitr::kable(digits=2, caption = "This sample is out of healthy range for these taxa.")



# mhg_outlier(mhg_range_for(tam.norm),mhg_range_for(subset_samples(people.norm, Site=="gut" & Condition == "Healthy"))) %>% filter(original>=max*0.8) %>% filter(original != 0)

# mhg_density_for_taxa(merge_phyloseq(gut.norm,people.norm), "Hespellia") +
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(people.norm, Username == "tm")
#   )["Hespellia"]) / 10000,
#   xend = as.numeric(otu_table(
#   subset_samples(people.norm, Username == "tm")
#   )["Hespellia"]) / 10000 ,
#   y = -Inf,
#   yend = Inf
#   )
# 
# mhg_density_for_taxa(merge_phyloseq(gut.norm,people.norm), "Parasporobacterium") +
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(people.norm, Username == "sj")
#   )["Parasporobacterium"]) / 10000,
#   xend = as.numeric(otu_table(
#   subset_samples(people.norm, Username == "sj")
#   )["Parasporobacterium"]) / 10000 ,
#   y = -Inf,
#   yend = Inf
#   )

# mhg_all_who_have(merge_phyloseq(gut.norm,people.norm), "Adlercreutzia") %>% arrange(desc(Abundance))
```

Then I computed which of my own taxa are out of range of the healthy people (Table \@ref(tab:caseStudyMeOutOfRange))

```{r caseStudyMeOutOfRange}

mhg_outlier(mhg_range_for(gut.norm),mhg_range_for(subset_samples(people.norm, Site=="gut" & Condition=="Healthy"))) %>% filter(original>max) %>% arrange(desc(original)) %>% knitr::kable(digits=2, caption = "Out of healthy range")

```


## Russians

Are there any microbes that are unique to people of Russian ancestry?

I have a few samples from two Russian immigrant sisters living in the United States, plus another sample from their mother. I'll compare them to a native North American with Russian ancestry

What might they have in common?
```{r caseStudyRussian1}


r <- subset_samples(people.gut.norm, Site == "gut" & ( stringr::str_detect(Username,"^rs")))
max(sample_data(ub17)$SSR)
r1 <- subset_samples(r, Username == "rs1")
  
ub17 <- subset_samples(people.gut.norm, Site == "gut" &  Username == "ub17")

ub17d <- subset_samples(ub17, SSR == max(sample_data(ub17)$SSR))


psmr::make_ord_for(phyloseq::merge_phyloseq(r,ub17))

#ub17 doesn't have the "european" microbe that I usually find associated with living for long periods in the UK
psmr::mhg_taxa_plot(ub17, "Anaeroplasma")

# taxa found in ub17 but not Russians 
psmr::unique_taxa(r,ub17)

# ub17 is out of range from the other Russians on these taxa
psmr::mhg_outlier(psmr::mhg_range_for(ub17),psmr::mhg_range_for(r)) %>%
  mutate_at(vars(-taxa),function(x) x/10000) %>%
  knitr::kable(digits = 3,  caption = "ub17 Outliers Compared to Russians")

```

Here's another one:

```{r}

# analysis of a person with three samples, including nose.

# jd <- subset_samples(people.norm, Username == "jd")
# jd_gut <- subset_samples(jd,Site == "gut")
# 
# 
# 
# psmr::mhg_outlier(psmr::mhg_range_for(subset_samples(people.norm, Site == "gut" & SSR == 357699)),
#                   psmr::mhg_range_for(people.healthy.gut.genus)) %>% knitr::kable()

# unique to 2017 sample
# psmr::unique_taxa(subset_samples(jd_gut, SSR == 357699),subset_samples(jd_gut, SSR == 5308) )

# Unique to 2014 sample
# psmr::unique_taxa(subset_samples(jd_gut, SSR == 5308), subset_samples(jd_gut, SSR == 357699) )
# 
# p.family.gut <- subset_samples(people.family, Site=="gut")
# 
# sink("junk.txt")
# ord <- ordinate(p.family.gut,method = "NMDS", distance = "bray")
# plot_ordination(p.family.gut,ord, color = "Condition", label = "SSR")
# sink()
# file.remove("junk.txt")

#psmr::mhg_range_for(subset_samples(people.norm, Site == "nose" & Username == "jd"))

# psmr::mhg_outlier(psmr::mhg_range_for(subset_samples(people.norm, Site == "nose" & Username == "jd")),
#                   psmr::mhg_range_for(subset_samples(people.norm, Site == "nose" & Username != "jd")))  %>% knitr::kable()
                    
```

