## Families

Recent research published in Nature^[ [https://www.nature.com/articles/s41598-018-37298-9](https://www.nature.com/articles/s41598-018-37298-9) ] seems to show that microbiomes are more similar between people with close social relationships than between people who are genetically similar. Is this true?

I compared the microbiome samples people have sent of themselves with their spouses and siblings.


## Mother and Daughter

Here's another example, this one from a friend who tested both herself and her mother.

Here's what's in the daughter, but not the mother:

```{r}
ub16 <- subset_samples(people.norm, Username == "ub16" & Reads > 10000 )
sample_data(ub16)$Label = "Daughter"
ub16.mom <- subset_samples(people.norm, Username == "ub16m")
sample_data(ub16.mom)$Label = "Mom"

ub16.species <- subset_samples(people.species, Username == "ub16" & Reads > 10000 )
sample_data(ub16.species)$Label = "Daughter"

ub16.mom.species <- subset_samples(people.species, Username == "ub16m")
sample_data(ub16.mom.species)$Label = "Mom"

mhg_plot_top_heat(merge_phyloseq(ub16, ub16.mom))
```



```{r}
psmr::unique_taxa(ub16,ub16.mom)[1]/10000
```

and the genus found in the mother but not the daughter

```{r}
psmr::unique_taxa(ub16.mom,ub16)[1]/10000
```

Here are species in the daughter but not the mom:

```{r}
psmr::unique_taxa(ub16.species,ub16.mom.species)[1]/10000
```

and vice versa:

```{r}
psmr::unique_taxa(ub16.mom.species,ub16.species)[1]/10000
```

As a bonus, we also have data from the brother, who has been living in Eastern Europe for several years. Here are the samples compared:


```{r casestudyFamilyRareif}

bird <- subset_samples(people.norm, grepl("ub1(6)+",sample_data(people.norm)$Username))
bird.phylum <- subset_samples(people.phylum, grepl("ub1(6)+",sample_data(people.phylum)$Username))
bird.species <- subset_samples(people.species, grepl("ub1(6)+",sample_data(people.species)$Username))

set.seed(100)
bird.rare <- rarefy_even_depth(bird)
bird.phylum.rare <- rarefy_even_depth(bird.phylum)
mhg_abundance(bird.rare)/10000
kable(mhg_abundance(bird.phylum.rare)/10000, col.names = c("sister-jun13","sister-may9","brother","mother"))

```

Because the sister's sample was of low quality (under 5000 reads), I rareifed all of these samples to the same depth before creating these tables. 


```{r }
# kable(mhg_abundance(bird)/10000, col.names = c("sister-jun13","sister-may9","brother","mother"))
# 
ub3 <- subset_samples(people.norm,Username == "ub3" & Site == "gut")
meg <- merge_phyloseq(ub3,subset_samples(bird,Username == "ub16b"))
# 
# kable(mhg_abundance(meg)/10000, col.names = c("ub3-May16","ub3-Jun16","Brother"))
# 
# ub3.species <- subset_samples(people.species,Username == "ub3" & Site == "gut")
# meg.species <- merge_phyloseq(ub3.species,subset_samples(bird.species,Username == "ub16b"))
# 
 kable(mhg_abundance(meg,top=10)/10000, col.names = c("ub3-May16","ub3-Jun16","Brother"))

```

The brother has much higher _Megamonas_ than I've seen in other samples. Is that significant?

```{r}
#psmr::unique_taxa(subset_samples(ub3.species, Date > "2016-06-01"),subset_samples(bird.species,Username == "ub16b"))[-2]/10000
```

### Husband and wife

Now let's look at a husband and wife:

```{r}
kt.family <- subset_samples(people.family, Username %in% c("ktp","kp"))

people.tmp <- people.family
sample_data(people.tmp)$Username <-ifelse(sample_data(people.tmp)$Username %in% c("ktp","kp"),"KP","People")
make_ord_for(subset_samples(people.tmp, Site == "gut"))

```

As expected of people who live together, both are quite similar relative to other people I've studied.

What can we see at the overall phylum level? (Table \@ref(tab:caseStudyFamilesHWTable))

```{r caseStudyFamilesHWTable}
kt.phylum <- subset_samples(people.phylum, Username %in% c("ktp","kp"))

kable(mhg_abundance(kt.phylum, top = 5)/10000, col.name = c("wife","husband"), digits = 2,
      caption = "Phylum differences between a husband and wife.")

```

and let's look at the differences between them on the genus level:

```{r}

people.tmp <- subset_samples(people.norm, Site == "gut")
kt.genus <- subset_samples(people.tmp, Username %in% c("ktp","kp"))
ktp <- subset_samples(people.tmp, Username == "ktp") #wife
kp <- subset_samples(people.tmp, Username == "kp") #husband

mhg_plot_top_heat(kt.genus)
```

Darker colors mean *lower* abundance of the microbes on this heat map. Let's look at which ones are unique to the couple:

```{r caseStudyFamilyWifeHusbandUnique}
kable(psmr::unique_taxa(ktp,kp)[1]/10000, col.names = c( "Wife (not husband)%"))
kable(psmr::unique_taxa(kp,ktp)[1]/10000, col.names = c( "Husband (not wife)%"))
```

That's just a start. Please let me know if you find anything else that's interesting.

```{r}
# We can compare the two for a particular genus like this (husband = red)
# mhg_density_for_taxa(people.tmp, "Clostridium") +
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "ktp")
#   )["Clostridium"]) / 10000,
#   xend = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "ktp")
#   )["Clostridium"]) / 10000,
#   y = 0,
#   yend = 0.5
#   ) +
#   annotate(
#   geom = "segment",
#   color = "red",
#   x = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "kp")
#   )["Clostridium"]) / 10000,
#   xend = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "kp")
#   )["Clostridium"]) / 10000,
#   y = 0,
#   yend = 0.5
#   )




```


```{r}
# mhg_density_for_taxa(people.tmp, "Subdoligranulum") +
#   annotate(
#   geom = "segment",
#   color = "blue",
#   x = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "kp")
#   )["Subdoligranulum"]) / 10000,
#   xend = as.numeric(otu_table(
#   subset_samples(people.tmp, Username == "kp")
#   )["Subdoligranulum"]) / 10000,
#   y = 0,
#   yend = 5
#   )
```

