## Ketogenic Diet {#caseStudyKeto}

THIS IS A VERY EARLY DRAFT

What happens if you eat a very low-carb, high fat diet ?

One study finds higher levels of _Akkermansia_ and _Parabacteroides_ reduce production of gamma-glutamylation and modulate hippocampal GABA/glutamate ratios.^[@olson_gut_2018 ]

Here is the microbiome of a person who tried Keto for several weeks.

```{r plotKeto, echo = FALSE, warning = FALSE, message = FALSE}

bp1.norm <- subset_samples(people.norm, Username = "bp1") #phyloseq_for_user("Orli_Kadoch",count.normalized = TRUE)

# mhg_plot_top_heat <- function(ps){
#   ps.top <- prune_taxa(names(sort(taxa_sums(ps),TRUE)[1:25]), ps)
#   plot_heatmap(ps.top,sample.order = "Date", sample.label = "Label")
# }

mhg_plot_top_heat(subset_samples(bp1.norm, SSR %in% c("17701","17944","39365"))) + ggtitle("Keto Diet Before/After") + theme(legend.position = "off")

```

Now let's compare two people. The red line is the person above, who tracked their microbiome while beginning the diet; the blue line is a person who has been on a ketogenic diet for six months or so. Interestingly, the genus _Parasutterella_ seems in unusually high abundance in both cases. Does it somehow relate to a low carb diet?



```{r keto2}
people.tmp <- subset_samples(people.norm, Site == "gut")
db1 <- subset_samples(people.tmp, Username == "db1")

# We can compare the two for a particular genus like this (husband = red)
mhg_density_for_taxa(merge_phyloseq(people.tmp,gut.norm), "Parasutterella") +
  annotate(
  geom = "segment",
  color = "blue",
  x = as.numeric(otu_table(db1)["Parasutterella"]) / 10000,
  xend = as.numeric(otu_table(db1)["Parasutterella"]) / 10000,
  y = 0,
  yend = 1.5
  ) +
  annotate(
  geom = "segment",
  color = "Red",
  x = as.numeric(otu_table(subset_samples(people.tmp, Username == "bp1"))["Parasutterella"]) / 10000,
  xend = as.numeric(otu_table(subset_samples(people.tmp, Username == "bp1"))["Parasutterella"]) / 10000,
  y = 0,
  yend = 1.5
  )


```

